Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-23 manuscript.Rmd
- Updated figures, added them to repository, and included them into the text (should move them to end of doc); - Went through first part of the discussion.

@@

@@

Figure XXX plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

@@

Applying this measure to all datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of intrapersonal change over time, with some items showing noteably high intrapersonal change compared to interpersonal variance. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

Intrapersonal change is often substantially less pronounced than interpersonal differences, accounting for less than 10 percent on average across all surveys. On numerous questions, like those on civil liberties, abortion, generalized trust, and civic duty, the systematic variance attributable to intrapersonal change is essentially zero. For these and other questions with minimal systematic variance attributable from intrapersonal change, it seems there is not enough cultural change during adulthood to attribute differences in personal culture mainly to experiences and social transitions. For these topics, the primary source of persona cultural differences appeasr to stem from early-life experiences.

More important than adjudicating past debates, however, the approach presented in this paper offers insights into new theoretical inquiries regarding when change or stability in personal culture predominates and its significance in explaining cultural variances within a population.

For instance, we detailed how the contribution that intrapersonal change can make in understanding cultural differences varies substantially by question and by individuals' education. It plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, intrapersonal change is more important in explaining cultural variation among people without a college compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, it suggests that college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions for graduates less influential relative to individuals without such a college experience.

@@

To this point, the theoretical debate in cultural sociology has focused on the existence of personal cultural change among adults and its relative importance. We believe the "existence" debate is theoretically misguided. Personal cultural change will always exist to some degree in an adult population, and no viable theoretical model expects just one process --stability or change-- to characterize a population. Our results reinforce that assertion, finding a mixture of interpersonal differences and intrapersonal change for all items. Even when we observe very low levels of systematic variance attributable to intrapersonal change for particular questions, this is potentially attributable to insufficient data, insufficient resolution on the underlying belief, or insufficient duration of observation.

A more reasonable theoretical debate pushed by previous work is the relative importance of these two processes for explaining cultural variation. While change might always be prevalent, its impact on social differences might be small or large, and our approach has been designed to clarify the contribution of intrapersonal change to such differences. It is difficult to draw a single conclusion on this question across a wide range of questions measured on different scales and over different durations of time, but the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on these items relative to the amount of difference we observe across people.

>>>>

```{r fig1, out.width = "400px", fig.cap="Proportions of explained variance in items of personal culture; based on predictions at participants' wave mid-point."}
knitr::include_graphics("fig/variation_share_lm_mdpnt.png")
```

@@

```{r fig2, out.width = "300px", fig.cap="Proportion of explained variance attributable to intrapersonal change; based on predictions at participants' wave mid-point; supressed intercept model; survey indicators not shown."}
knitr::include_graphics("fig/meta_no_int.png")
```

@@

Figure 3 plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people with a college degree than those without a college degree.

```{r img, out.width = "450px", fig.cap="Proportions of intrapersonal change of explained variance by college degree; based on predictions at participants' wave mid-point."}
knitr::include_graphics("fig/FIG_college2.png")
```

@@

Applying this measure to all datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of intrapersonal change over time, with some items showing noteably high intrapersonal change compared to interpersonal differences. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

Intrapersonal change is often substantially less pronounced than interpersonal differences, accounting for less than 10 percent on average across all surveys. On numerous questions, like those on civil liberties, abortion, generalized trust, and civic duty, the systematic variance explained by intrapersonal change is essentially zero. For these questions, it seems there is not enough cultural change during adulthood to warrant attributing differences in personal culture mainly to experiences and social transitions; instead, the primary source of the observed differences appears to stem from early-life experiences.

More important than adjudicating past debates, however, the approach presented in this paper offers offers the possibility to pose new theoretical questions regarding when change or stability in personal culture predominates and its significance in explaining cultural variances within a population. To examplify this potential, we detailed how the proportion in cultural differences that intrapersonal change can explain not only varies substantially by survey item but also by individual characteristics such as their education. Intrapersonal change plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, it is more important in explaining cultural variation among people without a college compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, it suggests that college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions for graduates less influential relative to individuals without such a college experience.

@@

So far, the debate in cultural sociology has centered around the presence and significance of cultural change in adults. Yet, arguing solely over its existence sets the wrong theoretical focus. In any adult population, some degree of personal cultural change is inevitable. Likewise, no viable theoretical perspective would expect just one process --stable differences or internal change between people-- to characterize the differences in personal cultural observed between them. Our results underscore this perspective, showing a combination of interpersonal differences and intrapersonal change across all items. [DELETED SENTECE, MISSPLACED]

A more reasonable theoretical debate pushed by previous work is the relative importance of these two processes for explaining cultural variation. It is difficult to draw a single conclusion on this question across a wide range of questions measured on different scales and over different durations of time, but the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on these items relative to the amount of difference we observe across people.
